---
title: "Project 1 “Насколько стара мидия”"
output: github_document
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE, warning=FALSE}
# Настройки R и knitr ----------------------------
# опции R
options(width = 70, scipen = 16, digits = 3) 
# опции чанков по умолчанию
knitr::opts_chunk$set(tidy = FALSE, # сохранять форматирование кода
               fig.width = 24, # ширина рисунков по умолчанию
               fig.height = 16)  # высота рисунков по умолчанию
```

##Introduction
Conducting descriptive statistics.

##Materials and methods

The calculations were carried out in the R environment using the pwr, data.table, dplyr, plyr and tidyr packages. The ggplot2, grid.extra and corrplot packages were used to plot the graphs.


## Task 1. Open datasets and merge them.
Calculation
```{r include = FALSE} 

#set packages, that will be needed
library(ggplot2)
library(data.table)
library(plyr)
library(dplyr)
library(tidyr)
library(readr)
library(gridExtra)
library(zoo)
library(corrplot)
library(RColorBrewer)

```

Open files (please write your dir)
```{r}
#path = "~/home/daria/Bioinf/statistics_r/data/Data/"

read_files <- function(link){
setwd(link)
file_list <- list.files(path=link)
dataset <- data.frame()
for (i in 1:length(file_list)){
temp_data <- fread(file_list[i], stringsAsFactors = F, encoding = "UTF-8")
dataset <- rbindlist(list(dataset, temp_data), use.names = T)
}
return(dataset)
}

# Use function to read files in our dir
mol_data <- read_files('/home/daria/Bioinf/statistics_r/data/Project_1/')
```

## Task 2. EDAnalise.

We need to look at the structure of the resulting dataset and identify various errors in the data set or collection.

```{r}
#checking structure of table
str(mol_data)
```

We can see that the Rings and Length columns have a non-numeric data type, and the Sex column has a long and inconvenient name.

```{r include = FALSE}
#rename Sex
mol_data <- rename(mol_data, Sex = `Sex (1 – male, 2 – female, 3 – uvenil)`)
## check if everything is numbers for rings
unique(mol_data$Rings)
#find the 'nine'
which(mol_data$Rings == 'nine')
#change value 
mol_data$Rings[3372] <- 9

#### check if everything is numbers for others
unique(mol_data$Length) #there is one NA, it`ll be renamed automatically

unique(mol_data$Sex) #check unique Sex values
which(mol_data$Sex == 'three') #change three
mol_data$Sex[618] <- 3
which(mol_data$Sex == 'one') #change three
mol_data$Sex[1374] <- 1 #changing 
which(mol_data$Sex == 'male')
mol_data$Sex[1399] <- 1 #changing 
unique(mol_data$Sex) #check to be sure, that we`ve changed everything

unique(mol_data$Diameter)
unique(mol_data$Height)
unique(mol_data$Whole_weight)
unique(mol_data$Shucked_weight) #there are too much, but we could see later that
#there are only numbers
unique(mol_data$Viscera_weight) #there are too much, but we could see later that
#there are only numbers
unique(mol_data$Shell_weight)

#we see that length and rings are not numeric, and do sex as factor
mol_data$Length <- as.numeric(mol_data$Length)
mol_data$Rings <- as.numeric(mol_data$Rings)
mol_data$Sex <- as.factor(mol_data$Sex)
```

We have Sex as factor with 3 levels and other parameters as numeric.
```{r}
#check structure again
str(mol_data)
```

Now we need to find and all NAs. 
```{r}
find_na <- data.frame(sum(is.na(mol_data)), sum(is.na(mol_data$Sex)), 
                      sum(is.na(mol_data$Rings)), sum(is.na(mol_data$Length)), 
                      sum(is.na(mol_data$Diameter)), sum(is.na(mol_data$Height)), 
                      sum(is.na(mol_data$Whole_weight)), sum(is.na(mol_data$Shucked_weight)),
                      sum(is.na(mol_data$Viscera_weight)), sum(is.na(mol_data$Shell_weight)))
```
We see that there are 21 NAs. We can replace them with mean, median, or mode, depending on the shape of the distribution. On the other hand, we have only 21 NA in more than 4100 observations, which constitute only 0.5% of the total sample. Thus I decided to omit them.
```{r include = FALSE}
#Delete all rows with NAs in them
#mol_data <- mol_data %>% na.omit(mol_data) - this is not working
sum(is.na(mol_data))
new_mol <- na.omit(mol_data)
```


###Now we can go further and examine the shape of the distribution and find outliers.
```{r include = TRUE}
#let`s do summary of all dataset
summary(new_mol)
```

Start with Rings parameter. 
```{r echo=TRUE, message=TRUE, warning=TRUE, paged.print=TRUE}

###R does not have a standard in-built function to calculate mode. So we create a user function to calculate mode of a data set in R. This function takes the vector as input and gives the mode value as output.
# Create the function.
getmode <- function(v) {
   uniqv <- unique(v)
   uniqv[which.max(tabulate(match(v, uniqv)))]
}

#find measures of central tendency
stat_rings <- data.frame(Mean = mean(new_mol$Rings),
                        Mode = getmode(new_mol$Rings),
                        Median = median(new_mol$Rings))
                 

qq_rings <- ggplot(new_mol, aes(sample = Rings)) +  #qq plot 
  stat_qq() +
  stat_qq_line(colour = "red") +
  labs(title="Normal Q-Q plot",
       y = "Normal data quantiles", x = "Normal theoretical quantiles") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))

density_rings <- ggplot(new_mol, aes(x = Rings)) + #density plot
  geom_density() + 
  #since the mode and median are the same, for simplicity, there is only the median
  geom_vline(aes(xintercept=median(Rings),         
                 color="Median"), linetype="dashed",
             size=1) +
  geom_vline(aes(xintercept=mean(Rings),
                 color="Mean"), linetype="dashed",
             size=1) +
  scale_color_manual(name = "statistics", values = c(Median = "blue", Mean = "red")) +
  labs(title="Rings density plot",
       y = "Rings", x = "Density") +
    theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))
  
box_rings <- ggplot(new_mol, aes(x = Rings)) +  #box plot
  geom_boxplot() +
  labs(title="Rings box-plot",
       y = "All molluscs", x = "Rings") +
theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))

plot_rings <- grid.arrange(density_rings, qq_rings,
                           box_rings, ncol = 2) 
```
The average value of the Rings parameter is 10, the mode and median are 9. The Q-Q plot clearly shows that the quantile points almost lie on the theoretical normal line. The Density plot shows almost normal distribution except right tail. "Rings box-plot" shows about 20 outliers and close to normal distribution. 
The shape of the distribution is asymmetric, unimodal, the distribution has a slight positive skew. However, since the mode and median of this distribution coincide, and the median does not differ much, the distribution is very close to normal.

``` {r}
#calculate sd and find outliers
sd_rings <- sd(new_mol$Rings) #sd = 3.22
outlier_values <- length(boxplot.stats(new_mol$Rings)$out) #278 outliers, which was calculated with 1.5IQR rule

#Calculate outliers using Hampel filter
lower_bound_rings <- median(new_mol$Rings) - 3 * mad(new_mol$Rings)
upper_bound_rings <- median(new_mol$Rings) + 3 * mad(new_mol$Rings)
outlier_rings_HF <- length(which(new_mol$Rings < lower_bound_rings | new_mol$Rings > upper_bound_rings)) #136 - outliers decreased two times with this method

```
The interquartile range (IQR) is three, the number of outliers calculated by the 1.5 IQR method is 278, using the Hampel filter - 136.

Do the same for Length.
```{r}
#find measures of central tendency
stat_Length <- data.frame(Mean = mean(new_mol$Length),
                        Mode = getmode(new_mol$Length),
                        Median = median(new_mol$Length))
                 

qq_Length <- ggplot(new_mol, aes(sample = Length)) +  #qq plot 
  stat_qq() +
  stat_qq_line(colour = "red") +
 labs(title="Normal Q-Q plot",
       y = "Normal data quantiles", x = "Normal theoretical quantiles") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))

density_Length <- ggplot(new_mol, aes(x = Length)) + #density plot
  geom_density() + 
  #since the mode and median are the same, for simplicity, there is only the median
  geom_vline(aes(xintercept=median(Length),         
                 color="Median"), linetype="dashed",
             size=1) +
  geom_vline(aes(xintercept=mean(Length),
                 color="Mean"), linetype="dashed",
             size=1) +
  geom_vline(aes(xintercept=getmode(Length),
                 color="Mode"), linetype="dashed",
             size=1) +
  scale_color_manual(name = "statistics", values = c(Median = "blue", Mean = "red", Mode = "green")) +
labs(title="Length density plot",
       y = "Length", x = "Density") +
    theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))
  
box_Length <- ggplot(new_mol, aes(y = Length)) +  #box plot
  geom_boxplot() +
labs(title="Length box-plot",
       y = "All molluscs", x = "Length") +
theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))
plot_Length <- grid.arrange(qq_Length, density_Length,
                           box_Length, ncol = 2)
```
The mean of the Length parameter is 0.524, the mode is 0.550 and the median is 0.545. The Q-Q plot clearly shows that the quantile points do not lie on the theoretical normal line with two deviating tails. The Density plot shows chi-square distribution distribution except left tail. "Rings box-plot" shows about some outliers. 
The shape of the distribution is asymmetric, unimodal, the distribution has a slight negative skew. It is chi-square distribution. 

``` {r}
#calculate sd and find outliers
summary(new_mol$Length)
sd_Length <- sd(new_mol$Length) #sd = 0.12
outlier_values_Length <- length(boxplot.stats(new_mol$Length)$out) #48 outliers, which was calculated with 1.5IQR rule

#Calculate outliers using Hampel filter
lower_bound_Length <- median(new_mol$Length) - 3 * mad(new_mol$Length)
upper_bound_Length <- median(new_mol$Length) + 3 * mad(new_mol$Length)
outlier_Length_HF <- length(which(new_mol$Length < lower_bound_Length | new_mol$Length > upper_bound_Length)) #35 outliers 

```
The interquartile range (IQR) is 0.165, the number of outliers calculated by the 1.5 IQR method is 48, using the Hampel filter - 35.


Do the same for Diameter.
```{r}
#find measures of central tendency
stat_Diameter <- data.frame(Mean = mean(new_mol$Diameter),
                        Mode = getmode(new_mol$Diameter),
                        Median = median(new_mol$Diameter))
                 

qq_Diameter <- ggplot(new_mol, aes(sample = Diameter)) +  #qq plot 
  stat_qq() +
  stat_qq_line(colour = "red") +
  labs(title="Normal Q-Q plot",
       y = "Normal data quantiles", x = "Normal theoretical quantiles") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))

density_Diameter <- ggplot(new_mol, aes(x = Diameter)) + #density plot
  geom_density() + 
  #since the mode and median are the same, for simplicity, there is only the median
  geom_vline(aes(xintercept=median(Diameter),         
                 color="Median"), linetype="dashed",
             size=1) +
  geom_vline(aes(xintercept=mean(Diameter),
                 color="Mean"), linetype="dashed",
             size=1) +
  geom_vline(aes(xintercept=getmode(Diameter),
                 color="Mode"), linetype="dashed",
             size=1) +
  scale_color_manual(name = " ", values = c(Median = "blue", Mean = "red", Mode = "green")) +
labs(title="Diameter density plot",
       y = "Diameter", x = "Density") +
    theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))
  
box_Diameter <- ggplot(new_mol, aes(y = Diameter)) +  #box plot
  geom_boxplot() +
labs(title="Diameter box-plot",
       y = "All molluscs", x = "Diameter") +
theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))

plot_Diameter <- grid.arrange(qq_Diameter, density_Diameter,
                           box_Diameter, ncol = 2)
```
The mean of the Diameter parameter is 0.408, the mode is 0.450 and the median is 0.425. The Q-Q plot clearly shows that the quantile points almost lie on the theoretical normal line with two deviating tails. The Density plot shows close to normal distribution with left tail. "Rings box-plot" shows about some outliers. 
The shape of the distribution is asymmetric, unimodal, the distribution has a slight negative skew.  

``` {r}
#calculate sd and find outliers
summary(new_mol$Diameter)
sd_Diameter <- sd(new_mol$Diameter) #sd = 0.0992
outlier_values_Diameter <- length(boxplot.stats(new_mol$Diameter)$out) #58 outliers, which was calculated with 1.5IQR rule

#Calculate outliers using Hampel filter
lower_bound_Diameter <- median(new_mol$Diameter) - 3 * mad(new_mol$Diameter)
upper_bound_Diameter <- median(new_mol$Diameter) + 3 * mad(new_mol$Diameter)
outlier_Diameter_HF <- length(which(new_mol$Diameter < lower_bound_Diameter | new_mol$Diameter > upper_bound_Diameter)) #39 outliers 
```
The interquartile range (IQR) is 0.125, the number of outliers calculated by the 1.5 IQR method is 58, using the Hampel filter - 39.

Do the same for Height.
```{r}
#find measures of central tendency
stat_Height <- data.frame(Mean = mean(new_mol$Height),
                        Mode = getmode(new_mol$Height),
                        Median = median(new_mol$Height))
                 

qq_Height <- ggplot(new_mol, aes(sample = Height)) +  #qq plot 
  stat_qq() +
  stat_qq_line(colour = "red") +
  labs(title="Normal Q-Q plot",
       y = "Normal data quantiles", x = "Normal theoretical quantiles") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))

density_Height <- ggplot(new_mol, aes(x = Height)) + #density plot
  geom_density() + 
  #since the mode and median are the same, for simplicity, there is only the median
  geom_vline(aes(xintercept=median(Height),         
                 color="Median"), linetype="dashed",
             size=1) +
  geom_vline(aes(xintercept=mean(Height),
                 color="Mean"), linetype="dashed",
             size=1) +
  geom_vline(aes(xintercept=getmode(Height),
                 color="Mode"), linetype="dashed",
             size=1) +
  scale_color_manual(name = " ", values = c(Median = "blue", Mean = "red", Mode = "green")) +
labs(title="Height density plot",
       y = "Height", x = "Density") +
    theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))
  
box_Height <- ggplot(new_mol, aes(y = Height)) +  #box plot
  geom_boxplot() +
labs(title="Height box-plot",
       y = "All molluscs", x = "Height") +
theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))

plot_Height <- grid.arrange(qq_Height, density_Height,
                           box_Height, ncol = 2)
```
The mean of the Height parameter is 0.408, the mode is 0.450 and the median is 0.425. The Q-Q plot clearly shows that the quantile points almost lie on the theoretical normal line without tails. The Density plot shows close to normal distribution with right tail. "Rings box-plot" shows about some outliers. 
The shape of the distribution is asymmetric, unimodal, the distribution has a slight positive skew, because of few outliers. If we omit them, it would be normal.

``` {r}
#calculate sd and find outliers
summary(new_mol$Height)
sd_Height <- sd(new_mol$Height) #sd = 0.0418
outlier_values_Height <- length(boxplot.stats(new_mol$Height)$out) #28 outliers, which was calculated with 1.5IQR rule

#Calculate outliers using Hampel filter
lower_bound_Height <- median(new_mol$Height) - 3 * mad(new_mol$Height)
upper_bound_Height <- median(new_mol$Height) + 3 * mad(new_mol$Height)
outlier_Height_HF <- length(which(new_mol$Height < lower_bound_Height | new_mol$Height > upper_bound_Height)) #14 outliers 

```
The interquartile range (IQR) is 0.050, the number of outliers calculated by the 1.5 IQR method is 28, using the Hampel filter - 14.

Do the same for Whole weight.

```{r}
#find measures of central tendency
stat_Whole_weight <- data.frame(Mean = mean(new_mol$Whole_weight),
                        Mode = getmode(new_mol$Whole_weight),
                        Median = median(new_mol$Whole_weight))
                 

qq_Whole_weight <- ggplot(new_mol, aes(sample = Whole_weight)) +  #qq plot 
  stat_qq() +
  stat_qq_line(colour = "red") +
  labs(title="Normal Q-Q plot",
       y = "Normal data quantiles", x = "Normal theoretical quantiles") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))

density_Whole_weight <- ggplot(new_mol, aes(x = Whole_weight)) + #density plot
  geom_density() + 
  
  geom_vline(aes(xintercept=median(Whole_weight),         
                 color="Median"), linetype="dashed",
             size=1) +
  geom_vline(aes(xintercept=mean(Whole_weight),
                 color="Mean"), linetype="dashed",
             size=1) +
  geom_vline(aes(xintercept=getmode(Whole_weight),
                 color="Mode"), linetype="dashed",
             size=1) +
  scale_color_manual(name = " ", values = c(Median = "blue", Mean = "red", Mode = "green")) +
labs(title="Whole weight density plot",
       y = "Whole weight", x = "Density") +
    theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))
  
box_Whole_weight <- ggplot(new_mol, aes(y = Whole_weight)) +  #box plot
  geom_boxplot() +
labs(title="Whole weight box-plot",
       y = "All molluscs", x = "Whole weight") +
theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))

plot_Whole_weight <- grid.arrange(qq_Whole_weight, density_Whole_weight,
                           box_Whole_weight, ncol = 2)
```
The mean of the Whole weight parameter is 0.829, the mode is 0.223 and the median is 0.801. The Q-Q plot clearly shows that the quantile points do not lie on the theoretical normal line with one deviating tail. The Density plot shows chi-square distribution with right tail. "Rings box-plot" shows about some outliers. 
The shape of the distribution is chi-squired, asymmetric, bimodal.

``` {r}
#calculate sd and find outliers
summary(new_mol$Whole_weight)
sd_Whole_weight <- sd(new_mol$Whole_weight) #sd = 0.49
outlier_values_Whole_weight <- length(boxplot.stats(new_mol$Whole_weight)$out) #29 outliers, which was calculated with 1.5IQR rule

#Calculate outliers using Hampel filter
lower_bound_Whole_weight <- median(new_mol$Whole_weight) - 3 * mad(new_mol$Whole_weight)
upper_bound_Whole_weight <- median(new_mol$Whole_weight) + 3 * mad(new_mol$Whole_weight)
outlier_Whole_weight_HF <- length(which(new_mol$м < lower_bound_Whole_weight | new_mol$Whole_weight > upper_bound_Whole_weight)) #0 outliers 

```
The interquartile range (IQR) is 0.712, the number of outliers calculated by the 1.5 IQR method is 29, using the Hampel filter - 0.

Do the same for Shucked_weight.
```{r}
#find measures of central tendency
stat_Shucked_weight <- data.frame(Mean = mean(new_mol$Shucked_weight),
                        Mode = getmode(new_mol$Shucked_weight),
                        Median = median(new_mol$Shucked_weight))
                 

qq_Shucked_weight <- ggplot(new_mol, aes(sample = Shucked_weight)) +  #qq plot 
  stat_qq() +
  stat_qq_line(colour = "red") +
  labs(title="Normal Q-Q plot",
       y = "Normal data quantiles", x = "Normal theoretical quantiles") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))

density_Shucked_weight <- ggplot(new_mol, aes(x = Shucked_weight)) + #density plot
  geom_density() + 
  
  geom_vline(aes(xintercept=median(Shucked_weight),         
                 color="Median"), linetype="dashed",
             size=1) +
  geom_vline(aes(xintercept=mean(Shucked_weight),
                 color="Mean"), linetype="dashed",
             size=1) +
  geom_vline(aes(xintercept=getmode(Shucked_weight),
                 color="Mode"), linetype="dashed",
             size=1) +
  scale_color_manual(name = " ", values = c(Median = "blue", Mean = "red", Mode = "green")) +
labs(title="Shucked weight density plot",
       y = "Shucked weight", x = "Density") +
    theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))
  
box_Shucked_weight <- ggplot(new_mol, aes(y = Shucked_weight)) +  #box plot
  geom_boxplot() +
labs(title="Shucked weight box-plot",
       y = "All molluscs", x = "Shucked weight") +
theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))

plot_Shucked_weight <- grid.arrange(qq_Shucked_weight, density_Shucked_weight,
                           box_Shucked_weight, ncol = 2)
```
The mean of the Whole weight parameter is 0.360, the mode is 0.175 and the median is 0.336. The Q-Q plot clearly shows that the quantile points do not lie on the theoretical normal line with rwo deviating tails. The Density plot shows chi-square distribution with right tail. "Rings box-plot" shows about some outliers. 
The shape of the distribution is chi-squired, asymmetric, unimodal.

``` {r}
#calculate sd and find outliers
summary(new_mol$Shucked_weight)
sd_Shucked_weight <- sd(new_mol$Shucked_weight) #sd = 0.222
outlier_values_Shucked_weight <- length(boxplot.stats(new_mol$Shucked_weight)$out) #48 outliers, which was calculated with 1.5IQR rule

#Calculate outliers using Hampel filter
lower_bound_Shucked_weight <- median(new_mol$Shucked_weight) - 3 * mad(new_mol$Shucked_weight)
upper_bound_Shucked_weight <- median(new_mol$Shucked_weight) + 3 * mad(new_mol$Shucked_weight)
outlier_Shucked_weight_HF <- length(which(new_mol$Shucked_weight < lower_bound_Shucked_weight | new_mol$Shucked_weight > upper_bound_Shucked_weight)) #34 outliers 

```
The interquartile range (IQR) is 0.316, the number of outliers calculated by the 1.5 IQR method is 48, using the Hampel filter - 34.

Do the same for Viscera weight.
```{r}
#find measures of central tendency
stat_Viscera_weight <- data.frame(Mean = mean(new_mol$Viscera_weight),
                        Mode = getmode(new_mol$Viscera_weight),
                        Median = median(new_mol$Viscera_weight))
                 

qq_Viscera_weight <- ggplot(new_mol, aes(sample = Viscera_weight)) +  #qq plot 
  stat_qq() +
  stat_qq_line(colour = "red") +
  labs(title="Normal Q-Q plot",
       y = "Normal data quantiles", x = "Normal theoretical quantiles") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))

density_Viscera_weight <- ggplot(new_mol, aes(x = Viscera_weight)) + #density plot
  geom_density() + 
  geom_vline(aes(xintercept=median(Viscera_weight),         
                 color="Median"), linetype="dashed",
             size=1) +
  geom_vline(aes(xintercept=mean(Viscera_weight),
                 color="Mean"), linetype="dashed",
             size=1) +
  geom_vline(aes(xintercept=getmode(Viscera_weight),
                 color="Mode"), linetype="dashed",
             size=1) +
  scale_color_manual(name = " ", values = c(Median = "blue", Mean = "red", Mode = "green")) +
labs(title="Viscera weight density plot",
       y = "Viscera weight", x = "Density") +
    theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))
  
box_Viscera_weight <- ggplot(new_mol, aes(y = Viscera_weight)) +  #box plot
  geom_boxplot() +
labs(title="Viscera weight box-plot",
       y = "All molluscs", x = "Viscera weight") +
theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))

plot_Viscera_weight <- grid.arrange(qq_Viscera_weight, density_Viscera_weight,
                           box_Viscera_weight, ncol = 2)
```
The mean of the Viscera weight parameter is 0.181, the mode is 0.172 and the median is 0.171. The Q-Q plot clearly shows that the quantile points do not lie on the theoretical normal line with rwo deviating tails. The Density plot shows chi-square distribution with right tail. "Rings box-plot" shows about some outliers. 
The shape of the distribution is chi-squired, asymmetric, unimodal.

``` {r}
#calculate sd and find outliers
summary(new_mol$Viscera_weight)
sd_Viscera_weight <- sd(new_mol$Viscera_weight) #sd = 0.11
outlier_values_Viscera_weight <- length(boxplot.stats(new_mol$Viscera_weight)$out) #26 outliers, which was calculated with 1.5IQR rule

#Calculate outliers using Hampel filter
lower_bound_Viscera_weight <- median(new_mol$Viscera_weight) - 3 * mad(new_mol$Viscera_weight)
upper_bound_Viscera_weight <- median(new_mol$Viscera_weight) + 3 * mad(new_mol$Viscera_weight)
outlier_Viscera_weight_HF <- length(which(new_mol$Viscera_weight < lower_bound_Viscera_weight | new_mol$Viscera_weight > upper_bound_Viscera_weight)) #13 outliers 

```
The interquartile range (IQR) is 0.159, the number of outliers calculated by the 1.5 IQR method is 26, using the Hampel filter - 13.

Do the same for Shell weight.
```{r}
#find measures of central tendency
stat_Shell_weight <- data.frame(Mean = mean(new_mol$Shell_weight),
                        Mode = getmode(new_mol$Shell_weight),
                        Median = median(new_mol$Shell_weight))
                 

qq_Shell_weight <- ggplot(new_mol, aes(sample = Shell_weight)) +  #qq plot 
  stat_qq() +
  stat_qq_line(colour = "red") +
  labs(title="Normal Q-Q plot",
       y = "Normal data quantiles", x = "Normal theoretical quantiles") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))

density_Shell_weight <- ggplot(new_mol, aes(x = Shell_weight)) + #density plot
  geom_density() + 
  
  geom_vline(aes(xintercept=median(Shell_weight),         
                 color="Median"), linetype="dashed",
             size=1) +
  geom_vline(aes(xintercept=mean(Shell_weight),
                 color="Mean"), linetype="dashed",
             size=1) +
  geom_vline(aes(xintercept=getmode(Shell_weight),
                 color="Mode"), linetype="dashed",
             size=1) +
  scale_color_manual(name = " ", values = c(Median = "blue", Mean = "red", Mode = "green")) +
labs(title="Shell weight density plot",
       y = "Shell weight", x = "Density") +
    theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))
  
box_Shell_weight <- ggplot(new_mol, aes(y = Shell_weight)) +  #box plot
  geom_boxplot() +
labs(title="Shell weight box-plot",
       y = "All molluscs", x = "Shell weight") +
theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))

plot_Shell_weight <- grid.arrange(qq_Shell_weight, density_Shell_weight,
                           box_Shell_weight, ncol = 2)
```
The mean of the Shell weight parameter is 0.239, the mode is 0.275 and the median is 0.171. The Q-Q plot clearly shows that the quantile points do not lie on the theoretical normal line with rwo deviating tails. The Density plot shows chi-square distribution with right tail. "Rings box-plot" shows about some outliers. 
The shape of the distribution is chi-squired, asymmetric, bimodal with positive skew.

``` {r}
#calculate sd and find outliers
summary(new_mol$Shell_weight)
sd_Shell_weight <- sd(new_mol$Shell_weight) #sd = 0.139
outlier_values_Shell_weight <- length(boxplot.stats(new_mol$Shell_weight)$out) #35 outliers, which was calculated with 1.5IQR rule

#Calculate outliers using Hampel filter
lower_bound_Shell_weight <- median(new_mol$Shell_weight) - 3 * mad(new_mol$Shell_weight)
upper_bound_Shell_weight <- median(new_mol$Shell_weight) + 3 * mad(new_mol$Shell_weight)
outlier_Shell_weight_HF <- length(which(new_mol$Shell_weight < lower_bound_Shell_weight | new_mol$Shell_weight > upper_bound_Shell_weight)) #19 outliers 

```
The interquartile range (IQR) is 0.199, the number of outliers calculated by the 1.5 IQR method is 35, using the Hampel filter - 19.


### Now we can draw a picture with correlation between different parameters. 
Since we comply with all the criteria for the applicability of the test, namely:
1. Observations are random and independent of each other, 
2. Samples are random and independent of each other, 
3. A large sample (> 30 observations in a group), 
I have used Pearson test.
```{r}
#let`s do correlation plot
new_mol$Sex <- as.numeric(new_mol$Sex) #for calculation correlation we need to use only numbers
cor_M = cor(new_mol, method = 'pearson')

#let`s compute the matrix of p-value
cor.mtest <- function(mat, ...) {
  mat <- as.matrix(mat)
  n <- ncol(mat)
  p.mat<- matrix(NA, n, n)
  diag(p.mat) <- 0
  for (i in 1:(n - 1)) {
    for (j in (i + 1):n) {
      tmp <- cor.test(mat[, i], mat[, j], ...)
      p.mat[i, j] <- p.mat[j, i] <- tmp$p.value
    }
  }
  colnames(p.mat) <- rownames(p.mat) <- colnames(mat)
  p.mat
}

# matrix of the p-value of the correlation
p.mat <- cor.mtest(new_mol)

# correlation plot
col <- colorRampPalette(c("#BB4444", "#EE9988", "#FFFFFF", "#77AADD", "#4477AA"))
corrplot(cor_M, method="color", col=col(200),  
         type="upper", order="original", 
         addCoef.col = "black", # Add coefficient of correlation
         tl.col="black", tl.srt=45, #Text label color and rotation
         number.cex=0.55,
         # Combine with significance
         p.mat = p.mat, sig.level = 0.01, insig = "blank", 
         # hide correlation coefficient on the principal diagonal
         diag=FALSE 
         )

```
We can see a very high correlation between length, diameter, different types of weight, which was expected. On the other hand, Sex is low correlated with the size of mollusks, and, which is not entirely expected, the number of rings also does not have a strong correlation with size. 

## Task 3. Mean and standart deviation for Lenght depended on Sex
```{r}
Sex_group <- new_mol %>% group_by(new_mol$Sex) %>% summarize_at(vars(Length), list(name = mean, sd))  

```

## Task 4. What percentage of molluscs has Height less than 0.165?
```{r}
sum(new_mol$Height < 0.165)/4176*100 #70,9%
```
## Task 5. What is the Length variable greater than 92% of all observations?
```{r}
quantile(new_mol$Length, 0.92) #0.67
```
## Task 6. Create a new variable Lenght_z_scores and store the Length variable into it after standardizing it.

```{r}
Lenght_z_scores <- (new_mol$Length - mean(new_mol$Length)) / sd(new_mol$Length)
```

## Task 7. Compare the diameter of the mollusks with the number of rings 5 and 15.
```{r}
#filter rings
Rings_comp <- new_mol %>% filter(Rings == 5 | Rings == 15) %>% select(Diameter, Rings)

ggplot(Rings_comp, aes(x = Rings, y = Diameter, group=Rings)) +
  geom_boxplot() +
  labs(title="Comparison of the diameter of the mollusks\n with the number of rings 5 and 15.") +
theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))
```
We can see that it is significant the difference between clams with 5 and 15 rings. 

Since we comply with all the criteria for the applicability of the test, namely:
1. Observations are random and independent of each other, 
2. Samples are random and independent of each other, 
3. A large sample (> 30 observations in a group), 
I have used t.test.
```{r}
t.test(Diameter ~ Rings, Rings_comp)

```
The diameter of mollusks with different numbers of rings varies significantly (t196 = -34, p < 0.05). 
Based on this result we may conclude the medians of these two distributions differ. The distribution of one population is shifted to the left or right of the other,” which implies different medians.

## Task 8. We are especially interested in the Diametr and Whole_weight variables. What can you say about them?

Since we comply with all the criteria for the applicability of the test, namely:
1. Observations are random and independent of each other, 
2. Samples are random and independent of each other, 
3. A large sample (> 30 observations in a group), 
I have used Pearson test.
```{r}
cor.test(new_mol$Diameter, new_mol$Whole_weight,  method = "pearson")
```
The correlation coefficient between Diameter and Whole_weight are 0.926 and the p-value is <0.0000000000000002, which indicates a strong positive correlation. 

## Task 9. While you were working with the data, you probably noticed something interesting in it. You can come up with a couple of hypotheses and test them. We'd love to get a fresh perspective on our data. 

Firstly, despite the fact that I used parametric tests to calculate the correlation - Pearson's test, since the distributions are not entirely normal, it would be more accurate to use nonparametric tests (for example, Spearman, Mann-Whitney or Kholmogorov-Smirnov). On the other hand, we have a fairly large sample.

```{r}
#for example recalculate 8th exercise
cor.test(new_mol$Diameter, new_mol$Whole_weight,  method = "spearman")
```
The correlation coefficient between Diameter and Whole_weight are 0.971 (Spearman method) and the p-value is <0.0000000000000002, which indicates a strong positive correlation. Previous test has shown 0.926 correlation coefficient (Pearson method). 

Secondly, one interesting hypothesis to test is whether there is a correlation between sex and size of mussels or the number of rings. But Pearson's correlation coefficient was low (-0.35 to -0.45) (calculations are in the point 2). 
```{r}
#plot box-plots 
ggplot(new_mol, aes(x = Sex, y = Diameter, group=Sex)) +
  geom_boxplot() +
  labs(title="Comparison of the diameter of the mollusks different sex") +
theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))
```
It is also interesting that the number of rings - a factor that actually indicates the age of the mussel - has a rather low Pearson correlation coefficient. But we see a significant difference between 5 and 15 ring clams. 
```{r}
#plot box-plots 
ggplot(new_mol, aes(x = Rings, y = Diameter, group=Rings)) +
  geom_boxplot() +
  labs(title="Comparison of the diameter of the mollusks\n with different numbers of Rings") +
theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))
```
We see that the correlation may not be linear, but logarithmic.

### Conclusion.
We conducted ED analysis and tested some hypotheses.